---
title: "Denver Maps"
format: gfm
---

To create maps in R, you usually start with a shape file. This shape file can be downloaded from a data source, or it can be loaded using one of many R packages.

In this doc, we'll be using two downloaded shape files from the [Denver Open Data Catalog](https://www.denvergov.org/opendata/ "Denver Open Data Catalog"), overlayed on top of base map layers obtained from [OpenStreetMap](https://wiki.openstreetmap.org/wiki/Main_Page "OpenStreetMap")

```{r}
# map data
 # loads shape files
# access OpenStreetMap data

# data manipulation
library(data.table)

# data visualization


# utils
library(fs)
library(lubridate)
library(purrr)
```

First, we'll use the `'osmdata'` package to load in a base layer Denver map. This is done be specifying a location object with `osmdata::opq()` , then loading in any specific [map features](https://wiki.openstreetmap.org/wiki/Map_Features "Map features - OpenStreetMap Wiki") as desired. For a basic map, we'll want two road layers, water, and parks.

```{r}
library(osmdata)

# load base denver data
xlim <- c(-105.06, -104.91)
ylim <- c(39.63, 39.80)
denver_coords <- c(xlim[1], ylim[1], xlim[2], ylim[2])
denver <- opq(denver_coords)

# separate primary and secondary roads for mapping purposes
highway_tags <- 
  c(
    "motorway",
    "trunk",
    "primary",
    "motorway_junction",
    "trunk_link",
    "primary_link",
    "motorway_link"
  )
street_tags <- available_tags("highway")
street_tags <- setdiff(street_tags, highway_tags)

# remove footpath tags from streets
footpath_tags <- c("footway", "path", "steps", "cycleway")
street_tags <- setdiff(street_tags, footpath_tags)

# specific tags for parks
park_tags <- 
  c(
    "disc_golf_course",
    "garden",
    "nature_reserve",
    "park",
    "pitch",
    "playground",
    "recreation_ground",
    "garden",
    "pitch"
  )

load_osm_feature <- function(key, value, location) {
  return(osmdata_sf(add_osm_feature(location, key = key, value = value)))
}

layer_tags <- 
  list(
    "highway" = highway_tags,
    "highway" = street_tags,
    "natural" = "water",
    "leisure" = park_tags
  )

layer_names <-
  c(
    "highways",
    "streets",
    "water",
    "parks"
  )

library(purrr)
layers <- 
  layer_tags %>% 
  imap(load_osm_feature, location = denver) %>% 
  set_names(layer_names)
```

Next, we load in our data from the city of Denver. We us the `sf` package to load the shape file.

```{r}
library(sf)

# read crime data shape file from Denver gov
crime_shape <- st_read("data/shapes/crime/crime.shp")

# read traffic accidents data shape file from Denver gov
accident_shape <- st_read("data/shapes/traffic_accidents/traffic_accidents.shp")
```

Then we use the `data.table` package to clean the data.

```{r}
library(data.table)

setDT(crime_shape)
setDT(accident_shape)

# filter to only crimes we want to look at
crimes_of_interest <- 
  c(
    "auto-theft", 
    "aggravated-assault", 
    "burglary", 
    "robbery", 
    "murder",
    "arson"
  )
filt_crimes <- crime_shape[crime_shape$OFFENSE_CA %in% crimes_of_interest, ]

# catch-all to remove unknown locations
filt_crimes <- filt_crimes[filt_crimes$GEO_LON < -100 & filt_crimes$GEO_LAT > 35, ]

# filter to recent years
library(lubrdiate)
YEAR_RECENT <- 2019
filt_crimes <- filt_crimes[year(filt_crimes$FIRST_OCCU) > YEAR_RECENT, ]
```

```{r}
# filter to recent traffic accidents
filt_accident <- accident_shape[year(accident_shape$FIRST_OCCU) > YEAR_RECENT, ]
```

Once our shape files are loaded, we can plot using `ggplot2`.

```{r}
ggplot() + 
  aes(geometry = geometry) +
  geom_sf(data = streets$osm_lines, size = 0.5, color = '#eedede') +
  geom_sf(data = highways$osm_lines, color = '#ff9999', size = 1) + 
  geom_sf(data = parks$osm_polygons, fill = '#94ba8e') +
  geom_sf(data = water$osm_multipolygons, fill = 'lightblue') + 
  geom_sf(data = filt_crimes, alpha = 0.01, size = 1, fill = "black") +
  geom_sf(data = filt_accident, alpha = 0.01, size = 1, fill = "red") +
  coord_sf(
    xlim = xlim, 
    ylim = ylim,
    expand = FALSE
  ) +
  theme_minimal()
```
